Dependently Typed Languages in Statix

### 2023-01-16 paper-evcs/paper.tex
Changes based on Aron's feedback

\keywords{Spoofax, Statix, Dependent Types} %TODO mandatory; please add comma-separated list of keywords

@@

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{abstract}
Static type systems can greatly enhance the quality of programs, but implementing a type checker that is both expressive and user-friendly is challenging and error-prone. The Statix meta-language (part of the Spoofax language workbench) aims to make this task easier by automatically deriving a type checker from a declarative specification of the type system. However, so far Statix has not been used to implement dependent types, an expressive class of type systems which require evaluation of terms during type checking. In this paper, we present an implementation of a simple dependently typed language in Statix, and discuss how to extend it with several common features such as inductive data types, universes, and inference of implicit arguments. While we encountered some challenges in the implementation, our conclusion is that Statix is already usable as a tool for implementing dependent types.
\end{abstract}

@@

Spoofax is a textual language workbench: \label{key}a collection of tools that enable the development of textual languages \cite{spoofax}. When working with the Spoofax workbench, the Statix meta-language can be used for the specification of static semantics. To provide these advantages to as many language developers as possible, Statix aims to cover a broad range of languages and type systems. However, no attempts have been made to express dependently typed languages in Statix.

Dependently typed languages are different from other languages, because they allow types to be parameterized by values. This allows types to express properties of values that cannot be expressed in a simple type system, such as the length of a list or the well-formedness of a binary search tree. This expressiveness also makes dependent type systems more complicated to implement. Especially, deciding equality of types requires evaluation of the terms they are parameterized by.

This goal of this paper is to investigate how well Statix is fit for the task of defining a simple dependently-typed language. We want to investigate whether typical features of dependently typed languages can be encoded concisely in Statix. The goal is not only to show that Statix can implement it, but to investigate whether implementing a dependent type checker is easier in Statix than in a general-purpose programming language.

In section \ref{sec:coc} of this paper we show that Statix is already usable as a tool for implementing dependent types. There were some challenges in implementing the dependent type system, which this paper also discusses. Finally, in section \ref{sec:extensions} we describe how to extend the language with several common features such as inductive data types, universes, and inference of implicit arguments.

@@

In this section, we will describe how to implement a dependently typed language in Statix. In section \ref{sec:coc-syntax} we will describe the syntax of the language, then in section \ref{sec:coc-scopes} we will describe how scope graphs are used to type check the language. Section \ref{sec:coc-dynsyms} describes the dynamic semantics of the language, and finally \ref{sec:coc-typecheck} how to type check the language. The implementation of the language is available on GitHub. 
\url{https://github.com/JonathanBrouwer/master-thesis/}

@@

The base language that has been implemented is the Calculus of Constructions \cite{Coquand_Huet_1988}, the language at the top of the lambda cube \cite{lambda_cube}. One extra feature was added that is not present in the Calculus of Constructions: let bindings. Let bindings could be desugared by substituting, but this may grow the program size exponentially, so having them in the language is useful. The abstract syntax of the language is available in figure \ref{fig:syntax}.

\begin{figure}[h]
\begin{lstlisting}
Type        : Expr
Let         : ID * Expr * Expr -> Expr
Var         : ID -> Expr
FnType      : ID * Expr * Expr -> Expr
FnConstruct : ID * Expr * Expr -> Expr
FnDestruct  : Expr * Expr -> Expr
\end{lstlisting}
\caption{The syntax for the base language. FnConstruct is a lambda function, FnDestruct is application of a lambda function.}
\label{fig:syntax}
\end{figure}

@@

\begin{lstlisting}
let f = \T: Type. \x: T. x;
f (_: Type -> Type) (\x: Type. x)
\end{lstlisting}
The AST of this program in Aterm format\cite{aterm} would be:
\begin{lstlisting}
Let(
  "f", 
  FnConstruct("T", Type(), FnConstruct("x", Var("T"), Var("x"))),
  FnDestruct(
	FnDestruct(Var("f"), FnConstruct("_", Type(), Type())),
	FnConstruct("x", Type(), Var("x"))
  )
)
\end{lstlisting}

@@

To type check the base language, we need to store information about the names that are in scope at each point in the program. There are two different cases, names that do not have a known value (only a type), such as function arguments, and names that do have a known value, such as let bindings.\footnote{In non-dependent languages there is no such distinction, but because we may need the value of a binding to compare types, this is needed in dependently typed languages.}

In Statix, all this information can be stored in a \emph{scope graph} \cite{scope_graphs}, which is a feature of Statix. It is a graph consisting of nodes for scopes, labeled edges for visibility relations, scoped declarations for a relation, and queries for references. We only use a single type of edge, called \verb|P| (parent) edges. It also only has a single relation, called \verb|name|. This name stores a \verb|NameEntry|, which can be either a \verb|NType|, which stores the type of a name, or a \verb|NSubst|, which stores a name that has been substituted with a value.

@@

A unique requirement for dependently typed languages is beta reduction during type checking, since types may require evaluation to compare. Beta reduction is the process of reduction a term to beta normal form, which is the state where no further beta reductions are possible \cite{tapl}. It works by matching a term of the form \verb|(\x. b) e|. and substituting \verb|x| in \verb|b| with \verb|e|. Beta reduction applies this rule anywhere in the term, whereas beta head-reduction only applies this rule at the head (outermost expression) of the term, and produces a term in beta-head normal form.

We implemented beta reduction using a Krivine abstract machine\cite{krivine}. The machine can head evaluate lambda expressions with a call-by-name semantics. It works by keeping a stack of all arguments that have not been applied yet. This turned out to be the more natural way of expressing this compared to substitution-based evaluation relation. We originally tried to implement the latter, which works fine for the base language. However, it runs into trouble when implementing inductive data types; more information about this will be in the full master's thesis\cite{thesis}. An additional benefit is that abstract machines are usually more efficient than substitution-based approaches.

@@

We need to define multiple predicates that will be used later for type checking. First, the primary predicate is \verb|betaReduceHead|, that takes a scoped expression and a stack of applications, and returns a head-normal expression. The scope acts as the environment from \cite{krivine}, using \verb|NSubst| to store substitutions. All rules for \verb|betaReduceHead| are given in figure \ref{fig:beta-head-reduce-rules}. We use the syntax $\bhr{\scope{s1}{e1}}{t}{\scope{s2}{e2}}$ to express \verb|betaReduceHead((s1, e1), t) == (s2, e2)|. Figure \ref{fig:beta-head-reduce-rules} contains the rules necessary for beta head reduction of the language. One predicate that is used for this is the \verb|rebuild| predicate, which takes a scoped expression and a list of arguments and converts it to an expression by adding \verb|FnDestruct|s.

@@

		% Let rule
		\inferrule{ 
			\bhr 
			{ \scope{ \co{sPutSubst}(s, n, (s, v))}{b} } 
			{ t }
			{ \scope{ s' } { e' } }
		}{
			\bhr
			{ \scope{ s }{ \co{Let}(n, v, b) } }
			{ t }
			{ \scope { s' } { e' } }
		}

		% Var rule - NSubst
		\inferrule{ 
			\co{sGetName}(s, n) = \co{NSubst}(se, e) 
			\\ \bhr
			{\scope{se}{e}}
			{t}
			{\scope{se'}{e'}} 
		}{
			\bhr
			{\scope{s}{\co{Var}(n)}}
			{t}
			{\scope{se'}{e'}} 
		}

		% Var rule - NType
		\inferrule{ 
			\co{sGetName}(s, n) = \co{NType}(t) 
		}{ 
			\bhr
			{\scope{s}{\co{Var}(n)}}
			{t}
			{\co{rebuild}(s, \co{Var}(n), t)} 
		}

		% FnType rule
		\inferrule{ 
		} { 
			\bhr
			{\scope{s}{\co{FnType}(n, a, b)}}
			{[]}
			{\scope{s}{\co{FnType}(n, a, b)}} 
		}

		% FnConstruct rule - No args
		\inferrule{ 
		} { 
			\bhr
			{\scope{s}{\co{FnConstruct}(n, a, b)}}
			{[]}
			{\scope{s}{\co{FnConstruct}(n, a, b)}} 
		}

		% FnConstruct rule - Args
		\inferrule{ 
			\bhr
			{\scope{\co{sPutSubst}(s, n, t)}{b}}
			{ts}
			{\scope{s'}{e'}} 
		} { 
			\bhr
			{\scope{s}{\co{FnConstruct}(n, a, b)}}
			{(t::ts)}
			{\scope{s'}{e'}} 
		}

@@

\begin{lstlisting}
typeOfExpr : scope * Expr -> Expr
\end{lstlisting}
We can then start defining type checking rules for the language. We introduce a number of judgements for typing and equality together with their counterparts in Statix.
\begin{enumerate}
	\item $\toe{s}{e}{t}$ is the same as \verb|typeOfExpr(s, e) == t|
	\item $\beq{\scope{s1}{e1}}{\scope{s2}{e2}}$ is the same as \verb|expectBetaEq((s1, e1), (s2, e2))|
	\item $\bhr{\scope{s1}{e1}}{t}{\scope{s2}{e2}}$ is the same as \verb|betaReduceHead((s1, e1)) == (s2, e2)| \\ (The same as in section \ref{sec:coc-dynsyms})
	\item $\bred{s1}{e1}{e2}$ is the same as \verb|betaReduce((s1, e1)) == e2|
	\item $\scope{sEmpty}{e}$ is the same as $e$ (empty scopes can be left out)
\end{enumerate}

@@

		% Let rule
		\inferrule{ 
			\toe{s}{v}{vt}
			\\ \toe{\co{sPutSubst}(s, n, (s, v))}{b}{t}
		}{
			\toe{s}{\co{Let}(n, v, b)}{t}
		}

@@

		% Var rule - NType
		\inferrule{ 
			\co{sGetName}(s, n) = \co{NType}(t) 
		}{ 
			\toe{s}{\co{Var}(n)}{t}
		}

		% Var rule - NSubst
		\inferrule{ 
			\co{sGetName}(s, n) = \co{NSubst}(se, e) 
			\\ \toe{se}{e}{t}
		}{
		 	\toe{s}{\co{Var}(n)}{t}
		}

		% FnType rule
		\inferrule{ 
			\toe{s}{a}{at}
			\\ \beq{at}{\co{Type}()} 
			\\ \bred{s}{a}{a'}
			\\\\  \toe{\co{sPutType}(s, n, a')}{b}{bt}
			\\ \beq{bt}{\co{Type}()}
		} { 
			\toe{s}{\co{FnType}(n, a, b)}{\co{Type}()}
		}

@@

		% FnConstruct rule
		\inferrule{ 
			\toe{s}{a}{at}
			\\ \beq{at}{\co{Type}()}
			\\ \bred{s}{a}{a'} 
			\\\\ \toe{\co{sPutType}(s, n, a')}{b}{bt}
		} { 
			\toe{s}{\co{FnConstruct}(n, a, b)}{\co{FnType}(n, a', bt)}
		}

		% FnDestruct rule
		\inferrule{ 
			\toe{s}{f}{ft}
			\\ \bhr
				{\scope{s}{ft}}
				{[]}
				{\scope{s'}{\co{FnType}(da, dt, db)}} 
			\\\\ \toe{s}{a}{at} 
			\\ \beq{at}{\scope{sf}{dt}}
			\\ \bred{\co{sPutSubst}(s', n, (s, arg))}{db}{db'}
		} { 
			\toe{s}{\co{FnDestruct}(f, a)}{db'}
	 	}

@@

\subsection{Avoiding variable capture}
One problem with the implementation presented in the previous sections is that it does not avoid variable capture. Variable capture happens when a term becomes bound because of a wrong substitution. For example, according to the inference rules in figure \ref{fig:type-check-rules} the type of \verb|\T : Type. \T : T. T| is \verb|T : Type -> T : T -> T|. This is not the correct type, and even worse, it is not possible to express the correct type without renaming! The problem is that there are multiple names, and there is no way to distinguish between them. This problem needs to be addressed in any name-based approach. It can be solved by using scopes to distinguish names. A full explanation of this will be in the master's thesis\cite{thesis}.

\section{Extensions}
\label{sec:extensions}
This section will discuss some further ideas that can be explored to build upon what has already been shown. These will be fully described in the master's thesis\cite{thesis} that this paper is based on, this is just a look into what is possible.

@@

\paragraph*{Semantic Code Completion}
Finally, we explored how semantic code completion presented by Pelsmaeker et al. \cite{codecompletion} applies to dependently typed languages. It works well, only showing completions that are semantically possible.

@@

\section{Related Work}
The implementation in this paper requires performing substitutions in types immediately, as types don't have a scope. Van Antwerpen et al. \cite[sect 2.5]{scopes_as_types} present an implementation of System F that does lazy substitutions, by using scopes as types. It would be interesting to see if this approach could also apply to the Calculus of Constructions, where types can contain terms.

Another interesting comparison is to see how implementing a dependently typed language in Statix differs from implementing it in a general purpose language. The pi-forall language\cite{pi_forall} is a good example of a language with a similar complexity to the language presented in this paper. In principle, the implementations are very similar. For example, the inference rules presented in \cite{pi_forall} are similar to the inference rules presented in figure \ref{fig:type-check-rules} from this paper. The primary difference is that they use a bidirectional type system, whereas this paper does not.

>>>>

\keywords{Spoofax, Statix, Dependent Types, Scope Graphs, Calculus of Constructions}

\supplementdetails[subcategory ={ Source Code }]{The implementation of the language}{https://doi.org/10.5281/zenodo.7541014}

@@

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{Ralf L\"{a}mmel, Peter D. Mosses, and Friedrich Steimann}
\EventNoEds{3}
\EventLongTitle{Eelco Visser Commemorative Symposium (EVCS 2023)}
\EventShortTitle{EVCS 2023}
\EventAcronym{EVCS}
\EventYear{2023}
\EventDate{April 5, 2023}
\EventLocation{Delft, The Netherlands}
\EventLogo{}
\SeriesVolume{109}
\ArticleNo{30}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{abstract}
Static type systems can greatly enhance the quality of programs, but implementing a type checker that is both expressive and user-friendly is challenging and error-prone. The Statix meta-language (part of the Spoofax language workbench) aims to make this task easier by automatically deriving a type checker from a declarative specification of a type system. However, so far Statix has not been used to implement dependent types, which is a class of type systems which require evaluation of terms during type checking. In this paper, we present an implementation of a simple dependently typed language in Statix, and discuss how to extend it with several common features such as inductive data types, universes, and inference of implicit arguments. While we encountered some challenges in the implementation, our conclusion is that Statix is already usable as a tool for implementing dependent types.
\end{abstract}

@@

Spoofax is a language workbench: \label{key}a collection of tools that enable the development of textual languages~\cite{spoofax}. When working with the Spoofax workbench, the Statix meta-language can be used for the specification of static semantics. It is a declarative language that uses inference rules to define the static semantics. From these rules a type checker and other editor tools can be derived. Statix aims to cover a broad range of languages and type systems. However, no attempts have been made to express dependently typed languages in Statix until now.

Dependently typed languages are different from other languages, because they allow types to be parameterized by values. This allows types to express properties of values that cannot be expressed in a simple type system, such as the length of a list or the well-formedness of a binary search tree. This expressiveness also makes dependent type systems more complicated to implement. Especially, deciding equality of types requires normalization of the terms they are parameterized by.

This goal of this paper is to investigate how well Statix is fit for the task of defining a simple dependently-typed language. We want to investigate whether typical features of dependently typed languages can be encoded concisely in Statix. The goal is not only to show that Statix can implement it, but also that the implementation is concise.

In section \ref{sec:coc} of this paper we show that Statix is already usable as a tool for implementing dependent types. There were some challenges in implementing the dependent type system, which this paper also discusses. Then, in section \ref{sec:extensions} we describe how to extend the language with several common features such as inductive data types, universes, and inference of implicit arguments.

@@

In this section, we will describe how to implement a dependently typed language in Statix. In section \ref{sec:coc-syntax} we will describe the syntax of the language, then in section \ref{sec:coc-scopes} we will describe how scope graphs are used to type check the language. Section \ref{sec:coc-dynsyms} describes the dynamic semantics of the language, and finally \ref{sec:coc-typecheck} how to type check the language.

@@

The base language that has been implemented is the Calculus of Constructions~\cite{Coquand_Huet_1988}. One extra feature was added that is not present in the Calculus of Constructions: let bindings. Let bindings could be desugared by substituting, but this may grow the program size exponentially, so having them in the language is useful. The concrete syntax of the language is available in figure \ref{fig:syntax}.

\begin{figure}[h]
\begin{lstlisting}
Expr.Let = "let" ID "=" Expr ";" Expr
Expr.Type = "Type"
Expr.Var = ID
Expr.FnType = ID ":" Expr "->" Expr {right}
Expr.FnConstruct = "\\" ID ":" Expr "." Expr
Expr.FnDestruct = Expr Expr {left}
\end{lstlisting}
\caption{The concrete syntax for the base language. $\co{FnConstruct}$ is a lambda function, $\co{FnDestruct}$ is application of a lambda function.}
\label{fig:syntax}
\end{figure}

@@

\begin{lstlisting}
let f = \T: Type. \x: T. x;
f (_: Type -> Type) (\x: Type. x)
\end{lstlisting}

@@

To type check the base language, we need to store information about the names that are in scope at each point in the program. There are two different cases, names that do not have a known value (only a type), which are names created by function arguments, and names that do have a known value, which are names created by let bindings.\footnote{In non-dependent languages there is no such distinction, but because we may need \emph{the value} of a binding to compare types, this is needed in dependently typed languages.}

In Statix, all this information can be stored in a \emph{scope graph}~\cite{scope_graphs}. It is a graph consisting of nodes for scopes, labeled edges for visibility relations, scoped declarations for a relation, and queries for references. We only use a single type of edge, called \verb|P| (parent) edges. It also only has a single relation, called \verb|name|. This name stores a \verb|NameEntry|, which can be either a \verb|NType|, which stores the type of a name, or a \verb|NSubst|, which stores a name that has been substituted with a value. The Statix definition of these concepts is given below: 
\begin{lstlisting}
constructors
    NameType : Expr -> NameEntry
    NameSubst : scope * Expr -> NameEntry
relations
    name : ID -> NameEntry
\end{lstlisting}

@@

A unique requirement for dependently typed languages is beta reduction during type checking, since types may require evaluation to compare. Beta reduction is the process of reduction a term to its beta normal form, which is the state where no further beta reductions are possible~\cite{tapl}. It works by matching a term of the form \verb|(\x. b) e|. and substituting \verb|x| in \verb|b| with \verb|e|. Beta reduction applies this rule anywhere in the term, whereas beta head-reduction only applies this rule at the head (outermost expression) of the term, and produces a term in beta-head normal form.

We implemented beta reduction using a Krivine abstract machine~\cite{krivine}. The machine can head evaluate lambda expressions with a call-by-name semantics. It works by keeping a stack of all arguments that have not been applied yet. This turned out to be the more natural way of expressing this compared to substitution-based evaluation relation. We originally tried to implement the latter, which works fine for the base language. However, it runs into trouble when implementing inductive data types; more information about this will be in the full master's thesis~\cite{thesis}. An additional benefit is that abstract machines are usually more efficient than substitution-based approaches.

@@

We need to define multiple predicates that will be used later for type checking. First, the primary predicate is \verb|betaReduceHead|, that takes a scoped expression and a stack of applications, and returns a head-normal expression. The scope acts as the environment from~\cite{krivine}, using \verb|NSubst| to store substitutions. All rules for \verb|betaReduceHead| are given in figure \ref{fig:beta-head-reduce-rules}. We use the syntax $\bhr{\scope{s_1}{e_1}}{t}{\scope{s_2}{e_2}}$ to express \verb|betaReduceHead((s1, e1), t) == (s2, e2)|. The \verb|t| in this definition is the argument stack of the Krivine machine. Figure \ref{fig:beta-head-reduce-rules} contains the rules necessary for beta head reduction of the language. One predicate that is used for this is the \verb|rebuild| predicate, which takes a scoped expression and the stack of arguments (of the Krivine machine state) and converts it to an expression by adding \verb|FnDestruct|s. Its signature is:
\begin{lstlisting}
rebuild : (scope * Expr) * list((scope * Expr)) -> (scope * Expr)
\end{lstlisting}

@@

		% Let rule
		\inferrule{ 
			\bhr 
			{ \scope{ \co{sPutSubst}(s, x, (s, e))}{b} } 
			{ t }
			{ \scope{ s' } { b' } }
		}{
			\bhr
			{ \scope{ s }{ \co{Let}(x, e, b) } }
			{ t }
			{ \scope { s' } { b' } }
		}

		% Var rule - NSubst
		\inferrule{ 
			\co{sGetName}(s, x) = \co{NSubst}(s_e, e) 
			\\ \bhr
			{\scope{s_e}{e}}
			{t}
			{\scope{s_{e'}}{e'}} 
		}{
			\bhr
			{\scope{s}{\co{Var}(x)}}
			{t}
			{\scope{s_{e'}}{e'}} 
		}

		% Var rule - NType
		\inferrule{ 
			\co{sGetName}(s, x) = \co{NType}(t) 
		}{ 
			\bhr
			{\scope{s}{\co{Var}(x)}}
			{t}
			{\co{rebuild}(s, \co{Var}(x), t)} 
		}

		% FnType rule
		\inferrule{ 
		} { 
			\bhr
			{\scope{s}{\co{FnType}(x, a, b)}}
			{[]}
			{\scope{s}{\co{FnType}(x, a, b)}} 
		}

		% FnConstruct rule - No args
		\inferrule{ 
		} { 
			\bhr
			{\scope{s}{\co{FnConstruct}(x, a, b)}}
			{[]}
			{\scope{s}{\co{FnConstruct}(x, a, b)}} 
		}

		% FnConstruct rule - Args
		\inferrule{ 
			\bhr
			{\scope{\co{sPutSubst}(s, x, t)}{b}}
			{ts}
			{\scope{s'}{e'}} 
		} { 
			\bhr
			{\scope{s}{\co{FnConstruct}(x, \_, b)}}
			{(t::ts)}
			{\scope{s'}{e'}} 
		}

@@

\begin{lstlisting}
typeOfExpr : scope * Expr -> Expr
\end{lstlisting}
We can then start defining type checking rules for the language. We introduce a number of judgements for typing and equality together with their counterparts in Statix.
\begin{enumerate}
	\item $\toe{s}{e}{t}$ is the same as \verb|typeOfExpr(s, e) == t|
	\item $\beq{\scope{s1}{e1}}{\scope{s2}{e2}}$ is the same as \verb|expectBetaEq((s1, e1), (s2, e2))|
	\item $\bhr{\scope{s1}{e1}}{t}{\scope{s2}{e2}}$ is the same as \verb|betaReduceHead((s1, e1), t) == (s2, e2)| \\ (The same as in section \ref{sec:coc-dynsyms})
	\item $\bred{s1}{e1}{e2}$ is the same as \verb|betaReduce((s1, e1)) == e2|
	\item $\scope{sEmpty}{e}$ is the same as $e$ (empty scopes can be left out)
\end{enumerate}

@@

		% Let rule
		\inferrule{ 
			\toe{s}{v}{t_e}
			\\ \toe{\co{sPutSubst}(s, x, (s, e))}{b}{t_b}
		}{
			\toe{s}{\co{Let}(x, e, b)}{t_b}
		}

@@

		% Var rule - NType
		\inferrule{ 
			\co{sGetName}(s, x) = \co{NType}(t) 
		}{ 
			\toe{s}{\co{Var}(x)}{t}
		}

		% Var rule - NSubst
		\inferrule{ 
			\co{sGetName}(s, x) = \co{NSubst}(s_e, e) 
			\\ \toe{s_e}{e}{t}
		}{
		 	\toe{s}{\co{Var}(x)}{t}
		}

		% FnType rule
		\inferrule{ 
			\toe{s}{a}{t_a}
			\\ \beq{t_a}{\co{Type}()} 
			\\ \bred{s}{a}{a'}
			\\\\  \toe{\co{sPutType}(s, x, a')}{b}{t_b}
			\\ \beq{t_b}{\co{Type}()}
		} { 
			\toe{s}{\co{FnType}(x, a, b)}{\co{Type}()}
		}

@@

		% FnConstruct rule
		\inferrule{ 
			\toe{s}{a}{t_a}
			\\ \beq{t_a}{\co{Type}()}
			\\ \bred{s}{a}{a'} 
			\\\\ \toe{\co{sPutType}(s, x, a')}{b}{t_b}
		} { 
			\toe{s}{\co{FnConstruct}(x, a, b)}{\co{FnType}(x, a', t_b)}
		}

		% FnDestruct rule
		\inferrule{ 
			\toe{s}{f}{t_f}
			\\ \bhr
				{\scope{s}{t_f}}
				{[]}
				{\scope{s_f}{\co{FnType}(x, t_{da}, t_b)}} 
			\\\\ \toe{s}{a}{t_a} 
			\\ \beq{t_a}{\scope{s_f}{t_{da}}}
			\\ \bred{\co{sPutSubst}(s_f, x, (s, a))}{t_b}{t_b'}
		} { 
			\toe{s}{\co{FnDestruct}(f, a)}{t_b'}
	 	}

@@

\subsection{Avoiding variable capture}
One problem with the implementation presented in the previous sections is that it does not avoid variable capture. Variable capture happens when a term becomes bound because of a wrong substitution. For example, according to the inference rules in figure \ref{fig:type-check-rules} the type of \verb|\T : Type. \T : T. T| is \verb|T : Type -> T : T -> T|. This is not the correct type, and even worse, it is not possible to express the correct type without renaming! The problem is that there are multiple names, and there is no way to distinguish between them. This problem needs to be addressed in any name-based approach. It can be solved by using scopes to distinguish names. A full explanation of this will be in the master's thesis~\cite{thesis}.

\section{Extensions}
\label{sec:extensions}
This section will discuss some further ideas that can be explored to build upon what has already been shown. These will be fully described in the master's thesis~\cite{thesis} that this paper is based on, this is just an exploration of what is possible.

@@

\paragraph*{Semantic Code Completion}
Finally, we explored how semantic code completion presented by Pelsmaeker et al.~\cite{codecompletion} applies to dependently typed languages. It works well, only showing completions that are semantically possible.

@@

\section{Related Work}
The implementation in this paper requires performing substitutions in types immediately, as types don't have a scope. Van Antwerpen et al.~\cite[sect 2.5]{scopes_as_types} present an implementation of System F that does lazy substitutions, by using scopes as types. It would be interesting to see if this approach could also apply to the Calculus of Constructions, where types can contain terms.

Another interesting comparison is to see how implementing a dependently typed language in Statix differs from implementing it in a general purpose language. The pi-forall language~\cite{pi_forall} is a good example of a language with a similar complexity to the language presented in this paper. In principle, the implementations are very similar. For example, the inference rules of pi-forall are similar to the inference rules presented in figure \ref{fig:type-check-rules} from this paper. The primary difference is that they use a bidirectional type system~\cite{bidirectional}, whereas this paper uses Statix' unification.
